---
title: "Prediction Assignment Writeup"
author: "Ramesh Thyagarajan"
date: "January 27, 2016"
output: html_document
---
## Human Activity Recognition
Human Activity Recognition - HAR - has emerged as a key research area in the last years and is gaining increasing attention by the pervasive computing research community (see picture below, that illustrates the increasing number of publications in HAR with wearable accelerometers), especially for the development of context-aware systems. There are many potential applications for HAR, like: elderly monitoring, life log systems for monitoring energy expenditure and for supporting weight-loss programs, and digital assistants for weight lifting exercises.

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

```{r}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(RColorBrewer)
library(rattle)


```

Reproducibility - Set seed
```{r}
set.seed(55555)
```
Data retrieval and exploration
```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# trainfile<-"./Data/pml-training.csv"
# testfile<-"./Data/pml-testing.csv"

#download.file(trainUrl,trainfile,method="curl")
#download.file(trainUrl,testfile,method="curl")


training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))

```
Data partioning into training(60%) & testing (40%)

```{r}
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ] 
myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```
### Data clensing
Removing near zero variance variables
```{r}

nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]

```
Removing the first column as its a running serial number
```{r}
myTraining<-myTraining[c(-1)]
```
Clean variables with more than 60% NA
```{r}
trg <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trg)) {
            if( length( grep(names(myTraining[i]), names(trg)[j]) ) == 1)  {
                trg <- trg[ , -j]
            }   
        } 
    }
}

myTraining <- trg


```
Trasforming data sets

```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])  # remove the classe column
myTesting <- myTesting[clean1]         # allow only variables in myTesting that are also in myTraining
testing <- testing[clean2]             # allow only variables in testing that are also in myTraining

dim(myTesting)

```

```{r}
dim(testing)
```


Coercing the data into same type

```{r}
 
 for (i in 1:length(testing) ) {
     for(j in 1:length(myTraining)) {
         if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
             class(testing[j]) <- class(myTraining[i])
         }      
     }      
 }

# Get the same class between testing and myTraining
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```
## Prediction with Decision Trees

```{r}
modelFitDT <- rpart(classe ~., data=myTraining, method = "class")
fancyRpartPlot(modelFitDT)
```


```{r}

predictDT <- predict(modelFitDT, myTesting, type="class")
cmDT<- confusionMatrix(predictDT,myTesting$classe)
cmDT
```
##Prediction with Random Forest

```{r}
modelFitRF <- randomForest(classe ~., data=myTraining)
predictRF <- predict(modelFitRF, myTesting, type="class")
cmRF <- confusionMatrix(predictRF, myTesting$classe)
cmRF
```

```{r}
plot(modelFitRF)
```


```{r}
plot(cmRF$table, col = cmDT$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmRF$overall['Accuracy'], 4)))

```

## Predicting Results using the Test data
Random Forest model gave an accuray of 99.89% on test dataset, which more than the Decision Tree. The expected out-of-sample error is 100 - 99.89 = 0.11%
  
```{r, echo=FALSE}
predictRF_on_Test <- predict(modelFitRF, testing, type = "class")
predictRF_on_Test

```



```{r}
 Write_files = function(x){
     n=length(x)
     for (i in 1:n){
       filename = paste0("problem_id_", i,".txt")
       write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
     }
 
 }
# #Write_files(predictRF_on_Test)
```

### Source of Data
Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 
